---
title: "Chapter 5"
author: "Mohammad Imangholiloo"
date: "18 November 2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# About the project  
**Welcome to my page**  

This is the learning diary for the course 'Introduction to Open Data Science 2020'. I would like to further develop my data science skills in R and get familiar with Github.  
*My GitHub repository is https://github.com/Imangholiloo/IODS-project*

```{r}
# Let's get the date and time
date()
```
alternatively, `print(Sys.time())`

## ------------------------------------------------------------------------------------  


# *Exercise 5:* **Dimensionality reduction techniques**  
<br>
Mohamamd Imangholiloo  

## *Part 1) Data wrangling*  
*Note: this was done to prepare for next week*  
 
```{r}
#Read the “Human development” file into R
hd <- read.csv("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/human_development.csv", stringsAsFactors = F)

#Read the “Gender inequality” file into R
gii <- read.csv("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/gender_inequality.csv", stringsAsFactors = F, na.strings = "..")
```  

For data decription and more info please visit:  
http://hdr.undp.org/en/content/human-development-index-hdi and http://hdr.undp.org/sites/default/files/hdr2015_technical_notes.pdf

```{r}
#strcutre of data  
str(hd)  

#dimension of file (row* column)  
dim(hd)
```  

```{r}
#summary of file
summary(hd)
```  

<br>
```{r}
#column names  
colnames(hd)  
```  
<br>
```{r}
#see head and tail of the data  
head(hd)  
tail(hd)  
```  

Let's do the same for gii dataset  
```{r}
#strcutre of data  
str(gii)  

#dimension of file (row* column)  
dim(gii)  

#summary of file
summary(gii)

#column names
colnames(gii)

#see head and tail of the data
head(gii)
tail(gii)
```


Task 4: rename the variables with (shorter) descriptive names  
```{r}
colnames(hd)
colnames(hd) <- c('HDI_Rank', 'Country', 'HDI', 'Life_Expec_Birth', 
                  'Expec_yr_Edu', 'Mean_yr_Edu', 'GNI_per_Cap', 
                  'GNI_per_Cap_Min_HDI_Rank')
colnames(hd)
```
So you can see that column names were shorten.  


Let's do the same fo gii dataset  

```{r}
colnames(gii)
colnames(gii) <- c('GII_Rank', 'Country', 'GII', 'Mortality_R', 
                  'Adolescent_Birth_R', 'Representation_Parliament',
                  'Sec_Edu_Fem','Sec_Edu_Mal',
                  "Lab_Forc_Particip_R_Fem",
                  "Lab_Forc_Particip_R_Mal" )
colnames(gii)
```
So you can see that column names were shorten.  




**Task 5.** Mutate the “Gender inequality” data and create two new variables  
```{r}
library("dplyr")
gii <- mutate(gii, edu_R = (Sec_Edu_Fem/Sec_Edu_Mal))
head(gii)


gii <- mutate(gii, Lab_Forc_R = (Lab_Forc_Particip_R_Fem/Lab_Forc_Particip_R_Mal))
head(gii)
```
# Extra coding: alternative to mutate function  
Instead of using mutate, this is also possible.
```{r}
gii$edu_R <- gii$Sec_Edu_Fem/gii$Sec_Edu_Mal
```

**Task 6:** Join together the two datasets using the variable Country as the identifier.  
**TIP**: we use inner join to keep only observations in both data sets.  
```{r}
hd_gii_joint <- inner_join(hd, gii, 
                           by = 'Country', suffix = c("_hd", "_gi"))

head(hd_gii_joint)
dim(hd_gii_joint)

setwd("./data")
write.csv(hd_gii_joint, "human.csv")
```